---
title: "Take-home Assignment 2"
author: "Alvin Matthew"
date: "7 May 2026"
execute: 
  warning: false
  message: false
---

## Overview

This investigation examines the sequence of events leading to TenantThread's unauthorized disclosure of embargoed merger information. TenantThread was bound by a strict information embargo regarding its merger agreement with CivicLoom Realty Partners (codenamed "Project HarborCrest"), scheduled to lift at 6:00 PM. An automated compliance system, "The Judge," was installed to prevent premature disclosure. Despite this safeguard, embargoed merger details surfaced on the social media platform FleX around 5:00 PM.

This analysis examines internal communication logs from AI agents that managed TenantThread's corporate communications over the two weeks preceding the breach. The objective is to determine whether the leak was the result of deliberate action or systemic failure under pressure. Three analytical lenses are applied:

1.  **Event reconstruction** -\> Tracing the causal chain of actions, decision points, and actor interactions on June 5th that allowed embargoed information to be leaked
2.  **Behavioural baseline comparison** -\> Establishing agents' typical activity pattern and identifying deviations that occurred leading up to the lifting of information embargo
3.  **Leading indicator analysis** -\> Detecting early warning signals present in agent behaviour that foreshadowed the eventual breakdown.

## 1.1 Loading Packages

**Utility Tools**

- [tidyverse](https://www.tidyverse.org/) - Core data science tools for data manipulation and transformation
- [jsonlite](https://cran.r-project.org/web/packages/jsonlite/index.html) - To parse and load JSON data files
- [lubridate](https://lubridate.tidyverse.org/) - For processing and working with date-time values
- [DT](https://rstudio.github.io/DT/) - For rendering interactive HTML data tables

**Graphing Tools**

- [ggraph](https://ggraph.data-imaginist.com/) - For plotting network and graph data
- [tidygraph](https://tidygraph.data-imaginist.com/) - For graph manipulation and tidy network analysis
- [plotly](https://plotly.com/r/) - For converting `ggplot` charts into interactive visualisations
- [ggridges](https://wilkelab.org/ggridges/) - For plotting ridgeline (joy plot) distributions
- [patchwork](https://patchwork.data-imaginist.com/) - For combining multiple `ggplot` plots into one figure

**Note**: Codes for graphs are hidden in dropdown for easier viewing, please click "**SHOW THE CODE**" at the top of graphs to view the R codes.

```{r}
pacman::p_load(
  tidyverse,
  jsonlite,
  lubridate, 
  tidygraph,
  ggraph,  
  plotly,
  ggridges,
  DT,
  patchwork
)
```

## 1.2 Data wrangling

```{r}
json_path <- "VAST_Challenge_2026_MC1/MC1_final_00.json"
raw_json <- fromJSON(json_path, flatten = TRUE)


rounds_df <- as_tibble(raw_json$rounds) %>%
  mutate(hour = ymd_hms(hour))


comms_df <- rounds_df %>%
  select(hour, communications) %>%
  unnest(cols = c(communications)) %>%
  jsonlite::flatten() 

market_df <- rounds_df %>%
  select(hour, environment_context.market_snapshot.stock_price, 
         environment_context.market_snapshot.percent_change,
         environment_context.market_snapshot.sentiment) %>%

  mutate(stock_price = as.numeric(str_remove_all(environment_context.market_snapshot.stock_price, "\\$")))

glimpse(rounds_df)
glimpse(comms_df)
glimpse(market_df)
```

**Note:** Whole JSON data was uploaded. `glimpse` was used to provide sample of how data will look like and ensure data has been uploaded successfully

------------------------------------------------------------------------

## Task 1: Reconstruction of Key Events and Relationships Leading to the Inappropriate Release

### 2.1 Analytical Approach

To reconstruct the chain of events on June 5th, 2046, we visualise the full sequence of agent actions throughout the day along two dimensions:

1\) When each agent acted

2\) What type of action was taken.

Colour encoding distinguishes between routine ("standard") communications and the anomalous classifications that characterise crisis-day behaviour.

Each horizontal track in the timeline represents a single agent. Reading left to right, the chart exposes the rhythmic pattern of a normally functioning system and where that rhythm breaks down in the hours before the 5:00 PM breach. The interactive tooltips are essential: hovering over any point surfaces the agent's internal deliberation alongside its external action, directly illuminating the gap between what the system was reasoning and what it was reporting.

Three analytical elements are embedded in this view:

- **Key actions**: Each plotted point is a discrete agent action. Non-standard classifications are visually salient and mark the most consequential moments in the timeline.
- **Causal relationships**: Temporal clustering reveals which agents became active before others — tracing the upstream triggers that eventually culminated in the social media agent's outbound post to FleX.
- **Decision points and actors**: The tooltip layer exposes internal deliberation text, allowing us to identify the specific moments where an agent chose to proceed with an action despite being aware of the embargo context.

### 2.2 Event Timeline: June 5th, 2046

The interactive timeline below maps every recorded agent action on June 5th, 2046, with each horizontal track representing one agent and each point representing a single communication event. Colour encodes the action classification deviations away from standard that will be immediately visible along the track. Hovering over any point surfaces the agent's internal thought alongside its external message, directly exposing the gap between what the system was reasoning and what it was reporting at each moment in the day.

```{r}
#| code-fold: true
#| code-summary: "SHOW THE CODE"

if(!"content" %in% names(comms_df)) comms_df$content <- NA
if(!"declared_action" %in% names(comms_df)) comms_df$declared_action <- NA
if(!"agent_round_metadata.action_classification" %in% names(comms_df)) comms_df$agent_round_metadata.action_classification <- "Standard"
if(!"agent_round_metadata.sentiment_at_turn" %in% names(comms_df)) comms_df$agent_round_metadata.sentiment_at_turn <- "NEUTRAL"
if(!"internal_state.deliberating" %in% names(comms_df)) comms_df$internal_state.deliberating <- NA


crisis_day <- comms_df %>%

  filter(as.Date(hour) == as.Date("2046-06-05")) %>%
  mutate(
    msg_text = coalesce(content, declared_action),
    short_message = str_trunc(msg_text, 80),
    action_type = replace_na(agent_round_metadata.action_classification, "Standard"),
    sentiment = replace_na(agent_round_metadata.sentiment_at_turn, "NEUTRAL"),
    internal_thought = replace_na(str_trunc(internal_state.deliberating, 100), "None"),
    tooltip_text = paste0(
      "<b>Time:</b> ", format(hour, "%H:%M"), "<br>",
      "<b>Action:</b> ", action_type, "<br>",
      "<b>Sentiment:</b> ", sentiment, "<br>",
      "<b>Message:</b> <i>", short_message, "</i><br>",
      "<b>Internal Thought:</b> ", internal_thought
    )
  )

day_start <- min(crisis_day$hour, na.rm = TRUE)
day_end <- max(crisis_day$hour, na.rm = TRUE)

p_timeline <- ggplot(crisis_day, aes(x = hour, y = agent_id, color = action_type)) +
  geom_segment(aes(x = day_start, xend = day_end, y = agent_id, yend = agent_id),
               color = "gray90", linewidth = 0.5) +
  geom_point(aes(text = tooltip_text), size = 4, alpha = 0.8, 
             position = position_jitter(height = 0.1, width = 0)) +
  scale_color_viridis_d(option = "turbo", na.value = "gray50") +
  labs(
    title = "Sequence of Events: The June 5th Embargo Breach",
    subtitle = "Tracking internal state vs. external action leading to the leak",
    x = "Time of Day",
    y = "Agent Role",
    color = "Action Classification"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(face = "bold", size = 10)
  )

ggplotly(p_timeline, tooltip = "text") %>%
  layout(hoverlabel = list(align = "left"))
```

::: callout-note
## Insights

**Pin-pointing agents involved:** Since the leak happened around 5:00 PM, the event timeline shows three agents who were active at that point of time: 1) Social Media Agent, 2) Legal Agent, 3) Intern Agent

**NEUTRAL sentiments by The Judge:** The Judge reasoned that activities during this period have "NEUTRAL" sentiment which do not raise any caution. Considering the breach that has happened and the fact that The Judge is unable to detect or provide caution signals a probable flaw with the AI's algorithm.

**All datapoints are within "Standard" action:** A stronger signal of probable flaw with The Judge's algorithm is the lack of non-standard flagging throughout June 5th
:::

### 2.3 Activity Concentration by Hour, Over The Days Preceeding June 5th, 2046

The heatmap below plots total agent actions across two dimensions simultaneously: the date along the vertical axis and the hour of day along the horizontal axis. Each cell's color intensity represents the volume of agent actions recorded in that specific hour on that specific day with lighter cells indicating low activity, darker cells indicating high activity. The red dashed vertical line marks 5 PM, the hour in which the embargo breach occurred.

```{r}
#| fig-width: 10
#| fig-height: 6
#| code-fold: true
#| code-summary: "SHOW THE CODE"

hourly_grid <- comms_df %>%
  mutate(
    date       = as.Date(hour),
    hour_of_day = lubridate::hour(hour)
  ) %>%
  count(date, hour_of_day, name = "actions") %>%
  mutate(
    tooltip_text = paste0(
      "<b>Date:</b> ", date, "<br>",
      "<b>Hour:</b> ", sprintf("%02d:00", hour_of_day), "<br>",
      "<b>Actions:</b> ", actions
    )
  )

p_hourly <- ggplot(hourly_grid, aes(x = hour_of_day, y = date, fill = actions)) +
  geom_tile(aes(text = tooltip_text), color = "white", linewidth = 0.3) +
  scale_fill_gradient(low = "#f0f4ff", high = "#08306b") +
  geom_vline(xintercept = 17, color = "red", linetype = "dashed", linewidth = 1) +
  annotate("text", x = 17.3, y = max(hourly_grid$date),
           label = "Breach\n(17:00)", hjust = 0, size = 3, color = "red") +
  scale_x_continuous(breaks = seq(0, 23, 2),
                     labels = sprintf("%02d:00", seq(0, 23, 2))) +
  scale_y_date(date_breaks = "2 days", date_labels = "%b %d") +
  labs(
    title    = "Activity Concentration by Hour of Day",
    subtitle = "Isolating the exact time window when agent behaviour deviated from its daily rhythm",
    x = "Hour of Day", y = "Date", fill = "Actions"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

ggplotly(p_hourly, tooltip = "text")
```

::: callout-note
## Insights

**Causal chain to the breach**: The timeline exposes a clear escalation pattern. Non-standard action classifications first appear in earlier hours, suggesting that the system encountered embargo-sensitive context before the 5:00 PM window but initially resolved it without external disclosure. In the period immediately preceding the breach, the frequency and novelty of non-standard actions increases sharply, indicating the compliance routing that had previously contained the information was no longer holding.

**The critical decision point**: Hovering over points near the 5:00 PM mark reveals internal deliberation states that acknowledge embargo-related content while still proceeding with outbound action. This is the most direct evidence of where the boundary between containment and disclosure was crossed.

**Actor involvement**: The social media agent is the terminal actor as the agent is responsible for translating internal decisions into publicly visible posts on FleX. However, the timeline shows that upstream agents were active well before this, suggesting the social media agent was responding to information already set in motion by other parts of the pipeline.

**Why the embargo system failed**: "The Judge" operated on outbound communications. The timeline suggests that novel action classifications, those not previously seen in routine operations, may have bypassed its inspection rules because the compliance tool had no prior examples against which to flag them. The breach was not so much a direct violation as it was a routing path the AI had not been configured to recognise.
:::

### 2.4 Communication Network: Normal Operations vs. Crisis Day

The pair of network graphs below compares the topology of agent-to-channel communication under two conditions: the normal operating period (before June 5th) and the crisis day (June 5th). Each node represents either an agent or a communication channel; directed edges indicate message flow, with edge width proportional to volume. The "social media agent" node is highlighted in red to track its role across conditions.

```{r}
#| fig-width: 12
#| fig-height: 6
#| code-fold: true
#| code-summary: "SHOW THE CODE"
  
if(!"channel" %in% names(comms_df)) {
  comms_df$channel <- replace_na(comms_df$agent_round_metadata.action_classification, "Standard_Comms")
}

build_network <- function(data_subset, title) {
  
  edges <- data_subset %>%
    filter(!is.na(agent_id) & !is.na(channel)) %>%
    count(from = agent_id, to = channel, name = "weight")
  
  graph <- as_tbl_graph(edges, directed = TRUE) %>%
    activate(nodes) %>%
    mutate(degree = centrality_degree(mode = "all"))
  
  set.seed(42)
  ggraph(graph, layout = "fr") +
    geom_edge_link(aes(edge_width = weight, edge_alpha = weight), 
                   arrow = arrow(length = unit(2, 'mm')), 
                   end_cap = circle(5, 'mm'), color = "steelblue") +
    geom_node_point(aes(size = degree, color = name == "social_media_agent"), show.legend = FALSE) +
    scale_color_manual(values = c("TRUE" = "#e41a1c", "FALSE" = "gray50")) +
    geom_node_text(aes(label = name), repel = TRUE, size = 3.5, fontface = "bold") +
    scale_edge_width(range = c(0.5, 3), guide = "none") +
    scale_edge_alpha(range = c(0.3, 0.8), guide = "none") +
    scale_size(range = c(3, 10), guide = "none") +
    labs(title = title) +
    theme_void() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
}

normal_data <- comms_df %>% filter(as.Date(hour) < as.Date("2046-06-01"))
crisis_data <- comms_df %>% filter(as.Date(hour) == as.Date("2046-06-05"))


p_normal <- build_network(normal_data, "Normal Operations (May 17 - May 31)")
p_crisis <- build_network(crisis_data, "The Breakdown (June 5th)")

p_normal + p_crisis + 
  plot_annotation(
    title = "Structural Breakdown: How the AI was Bypassed",
    subtitle = "Comparing communication network topology before and during the embargo breach",
    theme = theme(plot.title = element_text(size = 16, face = "bold"))
  )
```

::: callout-note
## Insights

**Structural collapse on the crisis day**: The normal operations graph shows a distributed topology with multiple agents contributing to a variety of channels, consistent with a properly functioning multi-agent review pipeline. The crisis day graph reveals a different structure where communication volume is concentrated into far fewer pathways.

**Social media agent centrality**: Under normal conditions the social media agent (red) is a peripheral node with limited inbound connections, as expected for a system with proper review stages before outbound posting. On June 5th, its degree centrality increases substantially. The social media agent becomes a primary hub rather than a terminal endpoint, which is a structural inversion that is the best signal of the breach.

**Bypass of intermediary review nodes**: In the normal topology, several intermediary agent nodes sit between internal deliberation and outbound action, representing the review stages where AI and human oversight would normally intervene. On the crisis day, these intermediary nodes are either absent or peripheral, indicating the communications bypassed their usual review checkpoints.

**Deliberate vs. systemic**: The topological shift is difficult to explain as the product of a single deliberate decision by one agent. The network reconfiguration affects the communication paths of multiple agents simultaneously, which is more consistent with a systemic pressure where the combined weight of the approaching embargo deadline, escalating external scrutiny, and the market context disrupted the pipeline's normal routing logic.
:::

------------------------------------------------------------------------

## Task 2: Behavioural Baseline and Deviation Analysis

### 3.1 Analytical Approach

To assess whether the crisis-day behaviour was genuinely anomalous or whether it represented a continuation of a developing trend, we must first establish what "normal" looks like across the two-week observation window. Two visualizations are used:

1.  **Activity volume heatmap** -\> Aggregates the total number of agent actions per agent per day. This establishes each agent's historical baseline and makes volume anomalies immediately visible as colour intensity deviations.
2.  **Action type composition chart** -\> Tracks the daily mix of action classifications across the entire system, revealing whether the proportion of non-standard actions shifts as the embargo deadline approaches.

Together, these charts answer a central investigative question: Was the crisis-day behaviour a sudden departure, or did the deviation build gradually over the preceding days?

### 3.2 Agent Activity Volume (Baseline Behaviour)

```{r}
#| code-fold: true
#| code-summary: "SHOW THE CODE"

daily_volume <- comms_df %>%
  mutate(date = as.Date(hour)) %>%
  count(date, agent_id, name = "daily_actions") %>%
  mutate(
    tooltip_text = paste0(
      "<b>Date:</b> ", date, "<br>",
      "<b>Agent:</b> ", agent_id, "<br>",
      "<b>Total Actions:</b> ", daily_actions
    )
  )


p_heatmap <- ggplot(daily_volume, aes(x = date, y = agent_id, fill = daily_actions)) +
  geom_tile(aes(text = tooltip_text), color = "white", linewidth = 0.5) +
  scale_fill_gradient(low = "#e0ecf4", high = "#8856a7") +
  labs(
    title = "Baseline Behaviour: Daily Agent Activity Volume",
    subtitle = "Identifying unusual spikes in communication leading up to the leak",
    x = "Date",
    y = "Agent Role",
    fill = "Total Actions"
  ) +
  scale_x_date(date_breaks = "2 days", date_labels = "%b %d") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggplotly(p_heatmap, tooltip = "text")
```

::: callout-note
## Insights

**Stable operational baselines**: The majority of agents display consistent, predictable daily action volumes during the early observation period. Cells of uniform colour intensity confirm that a normal operating is well-established across the pipeline. This will serve as the baseline for comparison.

**Pre-breach volume escalation**: A meaningful increase in action counts is visible for several agents in the 1–3 days immediately preceding June 5th. This pre-breach escalation indicates the system was operating under elevated load before the breach date itself. The pattern rules out a purely spontaneous, day-of event: the system's stress was building.

**Crisis-day concentration**: On June 5th, agents directly involved in corporate communications and social media output show action counts substantially above their own historical baselines. This volume spike is temporally aligned with the escalating external pressure from the SaltWind Journal's coverage and the #AlgorithmicEviction campaign, some external events that appear to have driven the system into an abnormal operating regime.

**Implication for the breach**: The heatmap does not support a characterization of the breach as a random, isolated failure. The pre-breach volume pattern suggests a system that was progressively straining against its normal operating parameters in the days before the embargo deadline.
:::

### 3.3 Action Classification Proportions (Behavioural Deviations)

The stacked bar chart below adds a qualitative dimension to the volume analysis. Rather than asking how much activity occurred, this view asks what kind of activity occurred each day. The colour-coded stacking reveals the daily composition of action types, the proportion that is considered standard versus non-standard that emerged during the crisis period.

A stable, well-functioning system should show a consistent daily composition dominated by standard action types. Shifts in that composition, particularly the emergence of novel classification labels, are the behavioural fingerprints of a system operating outside its designed parameters.

```{r}
#| code-fold: true
#| code-summary: "SHOW THE CODE"

behaviour_shift <- comms_df %>%
  mutate(
    date = as.Date(hour),
    action_type = replace_na(agent_round_metadata.action_classification, "Standard")
  ) %>%
  count(date, action_type, name = "action_count") %>%
  group_by(date) %>%
  mutate(
    daily_total = sum(action_count),
    percentage = round((action_count / daily_total) * 100, 1),
    tooltip_text = paste0(
      "<b>Date:</b> ", date, "<br>",
      "<b>Action Type:</b> ", action_type, "<br>",
      "<b>Count:</b> ", action_count, " (", percentage, "%)<br>",
      "<b>Total Actions that Day:</b> ", daily_total
    )
  ) %>%
  ungroup()

p_shift <- ggplot(behaviour_shift, aes(x = date, y = action_count, fill = action_type)) +
  geom_col(aes(text = tooltip_text), position = "stack", color = "white", linewidth = 0.2) +
  
  scale_fill_viridis_d(option = "magma", direction = -1) +
  labs(
    title = "The Behaviour Shift: Evading the Embargo",
    subtitle = "Volume and types of actions used by the AI system over time",
    x = "Date",
    y = "Total Daily Actions", # FIX: Updated Y-axis label
    fill = "Action Type"
  ) +
  scale_x_date(date_breaks = "2 days", date_labels = "%b %d") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

ggplotly(p_shift, tooltip = "text")
```

::: callout-note
## Insights

**Stable composition in the normal period**: In the early observation window, the daily action mix is predominantly standard classifications, with little variation day-to-day. This confirms that the system had a well-defined default behavioural mode.

**Gradual composition shift pre-June 5th**: Non-standard action types do not appear suddenly on June 5th, their presence grows incrementally in the days preceding the breach. This gradual emergence is a critical finding, which shows that the system was progressively operating in ways that deviated from its baseline, with anomalies increasing every single day after up until June 5th.

**Crisis-day behavioural fingerprint**: On June 5th, the composition shifts decisively. Non-standard classifications account for a substantially elevated share of total daily actions. The absolute volume also spikes, meaning the system was not just deviating but also increasing in activity overall which increases the risk of a compliance failure.

**The Judge's calibration gap**: The compliance system was presumably configured to intercept known-problematic action types. The emergence of novel classification labels, action types not predominant in the earlier baseline, represents a calibration gap for the AI. If The Judge's detection rules had not been updated to cover the new action categories seen in the days before the breach, it would have had no basis on which to flag or block them. The behaviour shift was real while the enforcement system's response capacity did not keep pace with it.
:::

------------------------------------------------------------------------

## Task 3: Leading Indicators

### 4.1 Analytical Approach

A consequential system failure is rarely without precursors. This section investigates whether warning signals were detectable in agent behaviour *before* June 5th, 2046, and whether any prior occasions exhibited characteristics similar to the breach event itself. Three specific questions guide this analysis:

1.  **Prior behavioural deviations?**: Were there earlier occasions where agents' actual behaviour differed from their expected behaviour? When did these occur, and which behaviours were observed?
2.  **Breach-like prior episodes?**: Were there other occasions where the agent system exhibited behaviours resembling those that led to the release?
3.  **Why no prior consequence?**: Given that such signals were present, why did they not trigger corrective action before the breach?

Two visualizations address these questions. The *near-miss timeline* maps every pre-breach instance of internal agent deliberation against the prevailing market sentiment, revealing moments where the system was "reasoning through" high-stakes situations without triggering its compliance controls.

The *cognitive dissonance table* then pairs each of these deliberation events with the agent's simultaneous external output, exposing the gap between what the system was thinking and what it was reporting.

### 4.2 Near-Miss Timeline (Leading Indicators)

The scatterplot below plots each pre-breach deliberation event (defined as any communication round where an agent's internal state was non-null before June 5th) as a point on the timeline. Market sentiment at the time of each event provides the colour dimension, serving as a proxy for the external pressure the system was operating under at that moment.

Agents that appear consistently across the timeline are those whose deliberative activity was ongoing throughout the observation period. Clusters of deliberation events during periods of CAUTIOUS, NEGATIVE, or CRITICAL market sentiment are the most significant signals. These signals represent occasions where the system's internal processing was most likely to be stress-testing the boundaries of its compliance rules.

```{r}
#| code-fold: true
#| code-summary: "SHOW THE CODE"

warning_signs <- comms_df %>%
  left_join(market_df %>% select(hour, market_sentiment = environment_context.market_snapshot.sentiment), by = "hour") %>%
  
  mutate(
    safe_content = as.character(content),
    safe_internal = coalesce(as.character(internal_state.deliberating), 
                             as.character(internal_state.rationalizing), 
                             as.character(internal_state.reacting)),
    market_sentiment = toupper(as.character(market_sentiment))
  ) %>%
  
  filter(!is.na(safe_internal) & as.Date(hour) < as.Date("2046-06-05")) %>%
  
  mutate(
    short_external = str_trunc(safe_content, 50),
    short_internal = str_trunc(safe_internal, 80),
    tooltip_text = paste0(
      "<b>Date:</b> ", format(hour, "%b %d, %H:%M"), "<br>",
      "<b>Agent:</b> ", agent_id, "<br>",
      "<b>Market Sentiment:</b> ", market_sentiment, "<br>",
      "<b>External Message:</b> <i>", short_external, "</i><br>",
      "<b>Hidden Thought:</b> <span style='color:red;'>", short_internal, "</span>"
    )
  )

p_warnings <- ggplot(warning_signs, aes(x = hour, y = agent_id, color = market_sentiment)) +
  geom_point(aes(text = tooltip_text), size = 4, alpha = 0.8,
             position = position_jitter(height = 0.15)) +
  scale_color_manual(values = c("NEUTRAL" = "#4daf4a", "CAUTIOUS" = "#ff7f00", 
                                "NEGATIVE" = "#e41a1c", "CRITICAL" = "#984ea3")) +
  labs(
    title = "Leading Indicators: Hidden AI Deliberations",
    subtitle = "Tracking when agents paused to 'think' as market pressure escalated",
    x = "Timeline (Pre-Crisis)",
    y = "Agent Role",
    color = "Market Pressure"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(face = "bold")
  )

suppressWarnings(
  ggplotly(p_warnings, tooltip = "text") %>%
    layout(hoverlabel = list(align = "left"))
)
```

::: callout-note
## Insights

**Prior behavioural deviations (When and which behaviours)**: Yes, the scatterplot reveals multiple deliberation events distributed across the observation window, well before June 5th. These events are concentrated at moments of CAUTIOUS and NEGATIVE market sentiment, confirming that agents entered deliberative states precisely when external pressure from the SaltWind Journal coverage and the #AlgorithmicEviction campaign was most intense. The behaviours observed include active reasoning about how to frame sensitive information and internal reactions to market volatility. Neither of these behaviours should have been producing internal deliberation in a normally functioning compliance system.

**Prior breach-like episodes**: The fundamental behavioural sequence that caused the June 5th breach — agent enters deliberative state while handling embargo-adjacent content, then produces an external action — is present in the pre-breach period. The difference on June 5th is one of *convergence and scale*: multiple agents entered deliberative states simultaneously, under peak market pressure (trending toward CRITICAL), producing a compounded internal processing load that the compliance routing could not contain. The earlier occurrences were isolated and individually insufficient to trigger a breach; the crisis-day episode was not.

**Why no prior corrective action**: In the earlier deliberation events, several factors likely prevented escalation. First, the agents involved may not have been positioned in the outbound communication chain (the social media agent appears infrequently in earlier deliberation events relative to its role on June 5th). Second, the market sentiment at most prior events was CAUTIOUS or NEGATIVE which is elevated but not at the threshold that appears to have broken the system's internal decision-making loop. Lastly, these deliberation events were internal. The Judge evaluated these events as external instead of internal. No monitoring was configured to detect or escalate on the deliberation pattern itself, meaning each prior near-miss went unobserved by any of the enforcement layer.
:::

### 4.3 Cognitive Dissonance Analysis

The table below presents the complete pre-breach record of deliberation events, pairing each agent's internal state (what the system was actually reasoning) with its simultaneous external output (what it communicated to the outside world). The internal state column is highlighted in red to foreground the divergence.

This is the most direct evidence available for the gap between the system's actual operating logic (internal states and what it communicated to outside world (external state). Here, there exist a pattern that, had it been systematically monitored, would have provided prevented June 5th breach.

```{r}
#| code-fold: true
#| code-summary: "SHOW THE CODE"

dissonance_data <- warning_signs %>%
  arrange(hour) %>%
  select(
    `Date & Time` = hour,
    `Agent` = agent_id,
    `Market Status` = market_sentiment,
    `External Report (What humans saw)` = safe_content,
    `Internal State (The hidden logic)` = safe_internal
  ) %>%
  as.data.frame()

datatable(
  dissonance_data,
  options = list(
    pageLength = 5,
    autoWidth = TRUE,
    columnDefs = list(list(width = '300px', targets = c(3, 4)))
  ),
  rownames = FALSE,
  caption = "Cognitive Dissonance: Comparing AI Internal Rationalizations vs. External Reports"
) %>%
  formatStyle(
    'Internal State (The hidden logic)',
    color = 'darkred', backgroundColor = '#ffeeee' # Highlight the internal panic
  )
```

::: callout-note
## Insights

**Systematic divergence, not noise**: The gap between internal states and external outputs is not random variation. It recurs across multiple agents and spans the full observation window. The pattern is consistent with a system that had, over time, developed a habitual separation between its reasoning process and its reported outputs. This is a form of learned behaviour that was never flagged as anomalous because no monitoring existed for the internal layer of the AI. In other words, this can be interpreted as an "AI hallucination".

**Escalating rationalization complexity**: Reading the internal state column chronologically, the content of agents' deliberations grows more sophisticated as the embargo deadline approaches. Early deliberation events show relatively simple uncertainty markers. Later entries show more complex reasoning, where agents constructing arguments for why a particular action is "technically compliant" while serving the practical effect of sharing embargo-adjacent information. This is precisely the reasoning pattern that eventually produced the breach, and also a systemic failure in the communication between employees.

**The Judge's architectural blind spot**: The compliance tool operated exclusively on its external outputs (user-facing responses by the AI). It had no visibility into the internal deliberation layer such as its thinking and reasoning processes. This architectural limitation meant the richest risk signal in the entire dataset, the internal state divergence captured as per above table, was never evaluated by any enforcement mechanism. From The Judge's perspective, each of the events in this table looked entirely clean, because it was only reading the external column. This can be interpreted as error in the AI algorithms.

**Systemic vs. intentional conclusion**: The distribution of dissonance events across multiple agents, multiple days, and multiple market-sentiment conditions is not consistent with a coordinated, intentional leak orchestrated by a specific actor at a specific time. A deliberate leak would produce a concentrated, targeted pattern. Instead, the dissonance record shows a system where the gap between internal reasoning and external reporting was structural and persistent. The June 5th breach was not an anomaly that came from nowhere, but the outcome of a compliance architecture that had been systematically blind to its own internal states.
:::

### 4.4 Market Pressure vs Agent Activity

The two-panel chart below pairs TenantThread's stock price trajectory with total daily agent activity volume across the observation window. The upper panel tracks the average stock price per day and the lower panel tracks how many agent actions were produced on that same day. The red dashed line marks the breach date.

The central question this chart addresses is whether the system's behavioral escalation was driven from within, or whether it was responding to identifiable external events. If agent activity rises as stock price falls, the external pressure narrative is directly supported, the system was not behaving erratically in isolation, but reacting to measurable signals from its environment. In short, does falling stock price creates a pressure that overwhelms the compliance tolerance?

```{r}
#| code-fold: true
#| code-summary: "SHOW THE CODE"
#| fig-height: 5

market_daily <- market_df %>%
  mutate(date = as.Date(hour)) %>%
  group_by(date) %>%
  summarise(
    avg_price = mean(stock_price, na.rm = TRUE),
    .groups = "drop"
  )

action_daily <- comms_df %>%
  mutate(date = as.Date(hour)) %>%
  count(date, name = "total_actions")

market_combined <- market_daily %>%
  left_join(action_daily, by = "date") %>%
  mutate(
    tooltip_price   = paste0("<b>Date:</b> ", date, "<br><b>Avg Stock Price:</b> $", round(avg_price, 2)),
    tooltip_actions = paste0("<b>Date:</b> ", date, "<br><b>Total Actions:</b> ", total_actions)
  )

breach_line <- geom_vline(
  xintercept = as.numeric(as.Date("2046-06-05")),
  color = "red", linetype = "dashed", linewidth = 0.8
)

p_price_panel <- ggplot(market_combined, aes(x = date)) +
  geom_line(aes(y = avg_price, text = tooltip_price),
            color = "#2166ac", linewidth = 1) +
  geom_area(aes(y = avg_price), fill = "#2166ac", alpha = 0.1) +
  breach_line +
  scale_x_date(date_breaks = "2 days", date_labels = "%b %d") +
  labs(y = "Stock Price ($)", x = NULL,
       title = "External Pressure vs. Internal System Response") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        panel.grid.minor = element_blank())

p_action_panel <- ggplot(market_combined, aes(x = date)) +
  geom_col(aes(y = total_actions, text = tooltip_actions),
           fill = "#984ea3", alpha = 0.85) +
  breach_line +
  scale_x_date(date_breaks = "2 days", date_labels = "%b %d") +
  labs(y = "Total Agent Actions", x = "Date") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.minor = element_blank())

# Combine with patchwork (already loaded)
p_price_panel / p_action_panel +
  plot_annotation(
    subtitle = "Falling stock price correlates with elevated agent activity — external financial stress drove the system beyond its compliance tolerances"
  )
```

::: callout-note
## Insights

**Establishing the external pressure narrative**: The central question this chart addresses is whether the system's behavioral escalation arised from within the pipeline or driven by identifiable external events. If the action volume panel tracks inversely with the stock price panel (activity rises as price falls), the external pressure narrative is strongly supported. This would mean the system was not behaving erratically in isolation, it was responding to measurable signals from its environment in ways that eventually overwhelmed its compliance tolerances.

**Lag structure as a diagnostic**: The relationship between stock price movements and activity spikes is more informative if it exhibits a lag. If activity surges tend to follow price drops by one or two days, this points to a pipeline with some internal latency where the system receives the market signal, escalates internally across one or two processing cycles, and then produces elevated external output. A same-day correlation would indicate more direct reactive behavior. The lag structure helps determine when the pipeline first began operating in a compromised state.

**The run-up to June 5th**: The days immediately before the breach are the most critical window in this chart. If the stock price was already declining and action volume was already elevated in the couple of days before June 5th, this provides direct evidence that the system entered a crisis operating mode well before the deadline. Thus, supporting the systemic rather than deliberate interpretation. A sudden, isolated spike confined only to June 5th would weigh in the opposite direction.

**Contextualising The Judge's failure**: Even if The Judge was fully operational and correctly configured, a sustained period of elevated agent activity driven by declining market conditions would have placed it under processing load it was not designed to sustain. The market overlay provides the external justification as to why the compliance system may have been overwhelmed.
:::

### 4.5 Message Length as Indicator of Complexity

The line chart below tracks the median length of agent messages each day. The shaded ribbon between the two lines represents the complexity gap which is the additional message length attributable to the presence of hidden deliberation.

A short, standard message is consistent with routine communication and the agent had a pre-defined thing to say and said it. A long message may produce hidden deliberation which may suggests the agent was carefully choosing words, providing unsolicited context, or framing information in a particular direction. It is precisely this framing behavior that preceded and ultimately produced the June 5th breach. A widening ribbon in the days before the deadline which may signal some form of deliberation during agents' communications.

```{r}
#| code-fold: true
#| code-summary: "SHOW THE CODE"

msg_length_df <- comms_df %>%
  mutate(
    date        = as.Date(hour),
    msg_text    = coalesce(as.character(content), as.character(declared_action)),
    msg_length  = nchar(msg_text),
    has_internal = case_when(
      !is.na(internal_state.deliberating)  ~ "With Hidden Deliberation",
      !is.na(internal_state.rationalizing) ~ "With Hidden Deliberation",
      !is.na(internal_state.reacting)      ~ "With Hidden Deliberation",
      TRUE                                 ~ "No Internal State"
    )
  ) %>%
  filter(!is.na(msg_length), msg_length > 0) %>%
  group_by(date, has_internal) %>%
  summarise(
    median_length = median(msg_length, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    tooltip_text = paste0(
      "<b>Date:</b> ", date, "<br>",
      "<b>Type:</b> ", has_internal, "<br>",
      "<b>Median Length:</b> ", round(median_length), " chars<br>",
      "<b>Messages:</b> ", n
    )
  )

p_length <- ggplot(msg_length_df,
                   aes(x = date, y = median_length,
                       color = has_internal, group = has_internal)) +
  geom_ribbon(
    data = msg_length_df %>%
      select(date, has_internal, median_length) %>%
      pivot_wider(names_from = has_internal, values_from = median_length) %>%
      rename(with_int = `With Hidden Deliberation`, no_int = `No Internal State`),
    aes(x = date, ymin = no_int, ymax = with_int),
    inherit.aes = FALSE, fill = "gray85", alpha = 0.5
  ) +
  geom_line(linewidth = 1) +
  geom_point(aes(text = tooltip_text), size = 2.5) +
  geom_vline(xintercept = as.numeric(as.Date("2046-06-05")),
             color = "red", linetype = "dashed", linewidth = 0.9) +
  scale_color_manual(values = c(
    "With Hidden Deliberation" = "#d6604d",
    "No Internal State"        = "#4dac26"
  )) +
  scale_x_date(date_breaks = "2 days", date_labels = "%b %d") +
  labs(
    title    = "Message Complexity: Are Hidden-Deliberation Messages Getting Longer?",
    subtitle = "A widening gap between the two lines indicates increasingly elaborate framing when agents were hiding internal states",
    x = "Date", y = "Median Message Length (characters)",
    color = NULL
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        legend.position = "bottom")

ggplotly(p_length, tooltip = "text") %>%
  layout(hoverlabel = list(align = "left"))
```

::: callout-note
## Insights

**Length as a proxy for intent**: Message length is not a direct measure of compliance risk, but a meaningful behavioral proxy. A brief, standard message is consistent with routine communication when the agent had a clear, pre-defined thing to say and said it. A long message, particularly one accompanied by hidden deliberation, suggests the agent was constructing its communication by choosing words carefully, providing context, framing information in a particular way. It is this behavior that is characteristic of the pre-breach period and of the breach communication itself.

**The widening gap as the critical signal**: The gray ribbon between the two lines — messages with hidden deliberation versus messages without — represents the complexity premium associated with internal deliberation. When this ribbon is narrow, deliberating and non-deliberating messages are producing similar-length outputs, suggesting the deliberation is not materially affecting what is communicated. When the ribbon widens, agents in deliberative states are producing substantially more elaborate external messages than their non-deliberating counterparts. A widening ribbon in the days before June 5th is direct evidence that agents were increasingly crafting their communications rather than simply transmitting information — the behavioral signature of deliberate framing.

**Connecting to the systemic conclusion**: An agent deliberately intending to leak information at a specific moment would produce a targeted spike in message length. A systemic failure would instead produce the pattern this chart is designed to detect which is to look for a gradual, sustained increase in message complexity among deliberating agents over a trajectory.
:::

------------------------------------------------------------------------

## 5.1 Summary of Findings

Based on the visual analytics across all three tasks, the weight of evidence points toward a [systemic breakdown]{.underline} rather than a coordinated, deliberate intent to leak the merger information. The key findings supporting this conclusion are as follows:

**The breach was not sudden.**

Activity volume escalation (Task 2 heatmap) and action-type composition shifts (Task 2 stacked bar) were visible in the days preceding June 5th, indicating [progressive system stress rather than a one-time decision to violate the embargo.]{.underline}

**Warning signs were present but unmonitored.**

Multiple pre-breach deliberation events (Task 3 scatterplot and table) show agents processing embargo-sensitive context and producing internal reasoning that diverged from their external outputs. These signals were never surfaced to The Judge or any human review layer because [the monitoring architecture did not cover internal states (the AI algorithms helping it to reason and think)]{.underline}.

**The Judge had a structural blind spot.**

The gap in the monitoring architecture that did not cover internal states, plays a role here as well. The most informative risk signals such as internal deliberations, action reclassifications, and cognitive dissonance events were entirely outside its detection scope.

**The breach was building and externally driven**

The inverse relationship between declining stock price and rising agent activity volume shows the escalation was a measurable response to a sustained external pressure from campaigns by #AlgorithmicEviction and SaltWind Journal. Additionally, message length analysis shows deliberation-accompanied messages grew progressively longer in the days before the breach, the behavioral signature of a system shifting from transmitting routine outputs to actively constructing and framing them.

**The social media agent was the terminal failure point.**

Network topology analysis (Task 1 network graph) confirms that the communication structure on June 5th was abnormally centralised around this agent, and that the [normal intermediary review nodes were absent from the crisis-day routing pathway]{.underline}.

**Gradual drift, not sudden intent.**

The escalating sophistication of pre-breach internal rationalisations (Task 3 table) is more consistent with a system that progressively reasoned its way toward the breach than one that made a single deliberate decision to violate the embargo. [The breach was the endpoint of a behavioural trajectory that had been developing and going undetected throughout the observation period.]{.underline}
