---
title: "Take-home Exercise 1"
author: "Alvin Matthew"
date: 24 May 2026
format: 
  revealjs:
    theme: default     
    transition: slide  
    slide-number: true 
    center: false
    smaller: false
    include-in-header:
      text: |
        <style>
        .reveal {
          font-size: 18px;
        }
        </style>
execute: 
  echo: true           
  eval: true           
  warning: false       
---

# Overview

**Setting the Scene**

- Insurance providers often face under-performing portfolios, driven by unclear causes of poor profitability and high market volatility.
- This project explores how visual analytics can help uncover hidden loss drivers, risk patterns and segmentation opportunities, and improved underwriting strategies.
- Using a dataset from [Mendeley Data](https://data.mendeley.com/datasets/sw4jmdb2sm/1?utm_source=chatgpt.com), this analysis focuses on numerical, categorical, and temporal variables.
- Using univariate (one variable) and bivariate (two variables) analyses
- Plots, graphs, data cleaning, and presentation slides will be done using RStudio using R coding in Quatro Document HTML format.
- Empty data cells are ignored based on the Missing Completely at Random (MCAR) mechanism as all variables can be considered completely independent of any observed values.
  - **Our Objective:** [Gain deeper insights into portfolio performance and identify key factors affecting profitability and volatility in our portfolios]{.underline}

```{r}
#| include: false
pacman::p_load(
  tidyverse,    
  treemapify,   
  colorspace,   
  hrbrthemes,   
  ggdist,       
  ggstatsplot,  
  plotly,
  corrplot,
  dplyr,
  ggplot2,
  forcats,
  scales
)
```

```{r}
#| include: false
df <- read_delim("Dataset of motor insurance portfolio.csv", 
                 delim = ";", 
                 locale = locale(decimal_mark = "."),
                 show_col_types = FALSE)

clean_df <- df %>%
  drop_na(vehicle_age, age_driving_licence, fuel_type, vehicle_value) %>%
  
  mutate(
    policy_type = as.factor(policy_type),
    policy_status = as.factor(policy_status),
    business_type = as.factor(business_type),
    payment_frequency = as.factor(payment_frequency),
    bonus_score = as.factor(bonus_score),
    fuel_type = as.factor(fuel_type),
    vehicle_brand = as.factor(vehicle_brand),
    municipality_type = as.factor(municipality_type),
    circulation_area = as.factor(circulation_area),
    year = as.factor(year),
    
    total_exposure = as.numeric(total_exposure),
    total_claims = as.numeric(total_claims),
    total_premium = as.numeric(total_premium),
    total_incurred = as.numeric(total_incurred)
  ) %>%
  
  mutate(
    loss_ratio = ifelse(total_premium > 0, total_incurred / total_premium, 0),
    claim_frequency = ifelse(total_exposure > 0, total_claims / total_exposure, 0),
    claim_severity = ifelse(total_claims > 0, total_incurred / total_claims, 0)
  ) %>%
  
  filter(
    total_premium > 0, 
    driver_age >= 18, 
    vehicle_age >= 0
  )

glimpse(clean_df)
```

## Distribution of Drivers' Ages in Our Portfolio

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

avg_age <- median(clean_df$driver_age, na.rm = TRUE)
p_age <- ggplot(clean_df, aes(x = driver_age)) +
  geom_histogram(binwidth = 4, fill = "steelblue", color = "white") +
  # Add the vertical line for the average
  geom_vline(xintercept = avg_age, color = "darkred", linetype = "dashed", linewidth = 1.2) +
  labs(
    title = "Distribution of Driver Ages in the Portfolio",
    subtitle = paste("The dashed red line represents the median age:", round(avg_age, 1), "years"),
    x = "Driver Age",
    y = "Count of Policies"
  ) +
  theme_minimal()

print(p_age)
```

- Most drivers in our portfolio are aged around 40 years old (based on mode of distribution)

- Median stands at 48 years old. This creates a distribution that is right-skewed meaning half of our clients are 48 years old and younger while the other half are older than 48 years.

## Claim Frequency by Driver Age Segment

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
clean_df %>%
  mutate(age_group = cut(driver_age, breaks = seq(18, 90, by = 10))) %>%
  group_by(age_group) %>%
  summarise(avg_claim_freq = mean(claim_frequency, na.rm = TRUE)) %>%
  drop_na() %>%
  ggplot(aes(x = age_group, y = avg_claim_freq)) +
  geom_col(fill = "darkseagreen") +
  labs(
    title = "Claim Frequency Volatility by Driver Age Group",
    x = "Driver Age Group",
    y = "Average Claim Frequency"
  ) +
  theme_minimal()
```

- Based on the graph where the youngest drivers cohort is on the left, the oldest on the right, it is evident that younger drivers tend to make more claims

## Relation Between Claim Severity & Age of Drivers Involved

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5

severity_by_driver_age <- clean_df %>%
  filter(total_claims > 0) %>% 
  group_by(driver_age) %>%
  summarise(
    avg_severity = mean(claim_severity, na.rm = TRUE),
    claim_count = n()
  ) %>%
  filter(claim_count > 30)

r_value <- cor(severity_by_driver_age$driver_age, severity_by_driver_age$avg_severity, use = "complete.obs")
r_text <- paste0("r = ", round(r_value, 3))


b2 <- ggplot(severity_by_driver_age, aes(x = driver_age, y = avg_severity)) + 
  geom_point(aes(size = claim_count), color = "darkorange", alpha = 0.7) + 
  geom_smooth(method = "lm", color = "darkblue", se = TRUE, show.legend = FALSE) +
  theme_ipsum() +
  labs(
    title = paste0("Volatility: Claim Severity vs Driver Age (", r_text, ")"),
    subtitle = paste("Pearson Correlation Coefficient:", round(r_value, 3)),
    x = "Driver Age (Years)",
    y = "Average Claim Severity ($)",
    size = "Number of Claims"
  )

ggplotly(b2, tooltip = c("x", "y", "size"))
```

- It is noted that drivers aged 80 is an outlier with an extremely high average claim severity

- Based on the weak positive linear correlation coefficient (r=0.286). This means older drivers tend to have higher claim severity but the difference compared to younger drivers is relatively small.

  \* Refer to Appendix for formulas used

## Distribution of Premium Paid by Policy Type

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

clean_df %>%
  filter(total_premium < quantile(total_premium, 0.99, na.rm = TRUE)) %>%
  mutate(policy_type = fct_reorder(policy_type, total_premium, .fun = median)) %>%
  ggplot(aes(x = policy_type, y = total_premium, fill = policy_type)) +
  
  geom_violin(trim = FALSE, alpha = 0.5, color = NA) +
  
  geom_boxplot(width = 0.15, fill = "white", color = "gray30", alpha = 0.8, outlier.shape = NA) +
  
  scale_fill_viridis_d(option = "mako", direction = -1) +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_ipsum(base_family = "sans") +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Premium Distribution by Policy Type",
    subtitle = "Violin shapes show density; inner boxplots show median and interquartile ranges",
    x = "Policy Type",
    y = "Total Annual Premium ($)"
  ) +
  stat_summary(
    fun = median, 
    geom = "text",
    aes(label = paste0("$", round(after_stat(y), 0))), 
    position = position_nudge(x = 0.15),
    hjust = 0,
    color = "black", 
    fontface = "bold",
    size = 4
  )
```

- Basic third-party only coverage (TP) has the cheapest premium at median price of \$161 with narrow deviations, meaning most clients are paying prices around the median price.

- Comprehensive without excess (COMP_N) has the most expensive premium median price of \$565 with much wider deviations, meaning even though the typical premium is highest, actual prices may vary substantially across clients.

  \* Refer to Appendix for descriptions of policy types

## Frequency of Claims by Policy Type

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

claim_dist_df <- clean_df %>%
  mutate(
    claim_bucket = case_when(
      total_claims == 0 ~ "0 Claims",
      total_claims == 1 ~ "1 Claim",
      total_claims == 2 ~ "2 Claims",
      total_claims >= 3 ~ "3+ Claims"
    ),

    claim_bucket = factor(claim_bucket, levels = rev(c("0 Claims", "1 Claim", "2 Claims", "3+ Claims")))
  ) %>%

  count(policy_type, claim_bucket) %>%
  group_by(policy_type) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()


safe_order <- claim_dist_df %>% 
  filter(claim_bucket == "0 Claims") %>% 
  arrange(pct) %>%
  pull(policy_type)

claim_dist_df <- claim_dist_df %>%
  mutate(policy_type = factor(policy_type, levels = safe_order))


ggplot(claim_dist_df, aes(x = policy_type, y = pct, fill = claim_bucket)) +
  geom_col(width = 0.65, color = "white", linewidth = 0.5) +
  geom_text(
    data = filter(claim_dist_df, pct > 0.02),
    aes(label = percent(pct, accuracy = 0.1)), 
    position = position_stack(vjust = 0.5), 
    color = "white", fontface = "bold", size = 3.5
  ) +
  scale_fill_manual(values = c(
    "3+ Claims" = "darkred", 
    "2 Claims"  = "darkorange", 
    "1 Claim"   = "steelblue", 
    "0 Claims"  = "gray60"
  )) +
  scale_y_continuous(labels = percent_format()) +
  coord_flip() +
  theme_ipsum(base_family = "sans") +
  labs(
    title = "Claim Count Distribution by Policy Type",
    subtitle = "Proportion of policies filing 0, 1, 2, or 3+ claims across different coverage packages",
    x = "Policy Type",
    y = "Percentage of Policies",
    fill = "Number of Claims Filed:"
  ) +
  theme(
    legend.position = "top", 
    legend.justification = "left",
    panel.grid.major.y = element_blank()
  )
```

- Despite paying the highest premium out of all the policy types, Comprehensive without excess (COMP_N) clients have the highest claim rates as well. It is also noted that COMP_N has as significantly higher number of 3+ claims than any other policy type. In fact it is almost 5 times more than the next highest number of 3+ claims.

## Composition of Portfolio by Vehicle Brands (Treemap)

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

brand_summary <- clean_df %>%
  count(vehicle_brand) %>%
  mutate(prop = n / sum(n)) %>%
  filter(prop > 0.01) # Keep only brands with > 1% market share for clarity

u1 <- ggplot(brand_summary, aes(area = n, fill = n, label = vehicle_brand)) +
  geom_treemap() +
  geom_treemap_text(colour = "white", place = "centre", grow = TRUE) +
  scale_fill_continuous_sequential(palette = "Blues 3") +
  theme_ipsum(base_family = "sans") +
  labs(title = "Portfolio Composition by Vehicle Brand",
       subtitle = "Displaying brands with >1% portfolio share",
       fill = "Policy Count")

print(u1)
```

- Most of the vehicles we covered are from budget brands such as Renault, Citroen, Peugeot, Volkswagen, Ford, Opel, and Seat

- The three luxury brands that are notable are Audi, Mercedes, and BMW

- This means that our cost of claims are leaning towards the cheaper end due to covering more budget brands than luxury brands

# Total Incurred Losses Claim Type

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6


incurred_losses_df <- clean_df %>%
  select(liability_property_incurred, liability_injury_incurred, property_incurred, 
         theft_incurred, fire_incurred, glass_incurred, legal_protection_incurred, occupants_incurred) %>%
  summarise(across(everything(), sum, na.rm = TRUE)) %>%
  pivot_longer(cols = everything(), names_to = "incurred_type", values_to = "total_amount") %>%

  mutate(
    incurred_type = str_replace(incurred_type, "_incurred", ""),
    incurred_type = str_replace_all(incurred_type, "_", " "),
    incurred_type = str_to_title(incurred_type),
    incurred_type = fct_reorder(incurred_type, total_amount)
  )

ggplot(incurred_losses_df, aes(x = incurred_type, y = total_amount)) +
  geom_col(fill = "steelblue", width = 0.7) +
  geom_text(aes(label = scales::dollar(total_amount, scale = 1e-6, suffix = "M")), 
            hjust = -0.1, color = "gray30", fontface = "bold") +
  coord_flip() +
  theme_ipsum(base_family = "sans") +
  scale_y_continuous(labels = scales::dollar_format(scale = 1e-6, suffix = "M"), 
                     expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Total Incurred Losses by Coverage Type",
    subtitle = "Financial severity breakdown (Displayed in Millions)",
    x = "Coverage / Claim Type",
    y = "Total Incurred Amount ($)"
  ) +
  theme(panel.grid.major.y = element_blank())
```

- Third-party injury, Third-party property damage, and First-party property damage make up the most of our claims losses.

- These three claim types create losses amount to \$65.57 million, which make up 89.33% of total losses.

## Frequency & Severity Matrix of Claim Types

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 8

library(ggrepel) 

risk_matrix_df <- clean_df %>%
  summarise(
    port_exposure = sum(total_exposure, na.rm = TRUE),
    
    Property_Freq = sum(property_claims) / port_exposure,
    Injury_Freq = sum(liability_injury_claims) / port_exposure,
    Theft_Freq = sum(theft_claims) / port_exposure,
    Fire_Freq = sum(fire_claims) / port_exposure,
    Glass_Freq = sum(glass_claims) / port_exposure,
    
    Property_Sev = sum(property_incurred) / sum(property_claims),
    Injury_Sev = sum(liability_injury_incurred) / sum(liability_injury_claims),
    Theft_Sev = sum(theft_incurred) / sum(theft_claims),
    Fire_Sev = sum(fire_incurred) / sum(fire_claims),
    Glass_Sev = sum(glass_incurred) / sum(glass_claims)
  ) %>%
  pivot_longer(cols = -port_exposure, names_to = c("Coverage", ".value"), names_sep = "_") %>%
  
  mutate(
    label_text = paste0(
      Coverage, "\n(", 
      scales::dollar(Sev, accuracy = 1), ", ", 
      scales::percent(Freq, accuracy = 0.01), ")"
    )
  )


ggplot(risk_matrix_df, aes(x = Freq, y = Sev, label = label_text)) +
  geom_point(color = "darkred", size = 4) +
  

  geom_text_repel(fontface = "bold", size = 4.5, box.padding = 0.8, lineheight = 0.9, color = "midnightblue") +
  
  geom_vline(xintercept = median(risk_matrix_df$Freq), linetype = "dashed", color = "gray50") +
  geom_hline(yintercept = median(risk_matrix_df$Sev), linetype = "dashed", color = "gray50") +
  theme_ipsum(base_family = "sans") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Portfolio Risk Matrix: Frequency vs. Severity",
    subtitle = "Labels display: Coverage Type ($ Average Severity, % Annual Frequency)",
    x = "Claim Frequency (% Chance per Year of Exposure)",
    y = "Average Claim Severity ($)"
  )
```

- High Severity, Low Frequency: [Fire and Injury]{.underline}

- Low Severity, High Frequency: [Windshield/Glass and Property]{.underline}

- Low Severity, Low Frequency: Theft

  \* Refer to Appendix for formulas used

# Recommendations

- Based on the correlation between age and claim severity, adverse claim behaviours by COMP_N clients, and loss drivers due to property damage and injury

- Overhaul "COMP_N" underwriting rules to introduce dynamic pricing or introduce mandatory excess trigger that activates automatically after the second claim within a single policy year.

- Refine pricing models based on age because older drivers pose more severe financial volatility than younger cohorts. Implementing a cap on claims payout or requiring mandatory health checkups or even outright not willing to insure after a certain age may help stem our losses

- The portfolio suffers from high frequency of low severity claims such as property and glass claims. It is suggested to have a specialised repair networks and glass replacement vendors to process claims in bulk and reduce average costs.

# Appendix

[**Formulas used**]{.underline}

1.  $\text{Loss Ratio} = \frac{\text{Total Incurred Losses}}{\text{Total Premium}}$

2.  $\text{Claim Frequency} = \frac{\text{Total Claims}}{\text{Total Exposure}}$

3.  $\text{Claim Severity} = \frac{\text{Total Incurred}}{\text{Total Claims}}$

[**Policy Type Descriptions:**]{.underline}

- **TP** -\> Basic third-party liability coverage only

- **TPG** -\> Third-party liability with glass coverage

- **CC** -\> Third-party liability combined with two or more additional coverages (theft, fire, property or glass)

- **COMP_E** -\> Comprehensive without excess

- **COMP_N** -\> Comprehensive without excess
